### 2024-10-10 manuscript.tex
improve introduction and add citations [ci skip]
This build is based on
https://github.com/pivlab/annual_review_of_biomedical_data_science/commit/517d43c379923fb79c0ec18f4a2aa10ed2beb727.

This commit was created by the following CI build and job:
https://github.com/pivlab/annual_review_of_biomedical_data_science/commit/517d43c379923fb79c0ec18f4a2aa10ed2beb727/checks
https://github.com/pivlab/annual_review_of_biomedical_data_science/actions/runs/11282697412

This manuscript
(\href{https://pivlab.github.io/annual_review_of_biomedical_data_science/v/c213a22b1a038348f3a47b835a50d5d5e35f8e7b/}{permalink})
was automatically generated
from \href{https://github.com/pivlab/annual_review_of_biomedical_data_science/tree/c213a22b1a038348f3a47b835a50d5d5e35f8e7b}{pivlab/annual\_review\_of\_biomedical\_data\_science@c213a22}
on October 10, 2024.

@@

Understanding the genetics of complex traits remains a formidable challenge.
Genome-Wide Association Studies (GWAS) have been instrumental in advancing our understanding of the genetics behind complex traits.
By scanning the genomes of large populations, GWAS identify associations between genetic variants and traits, providing valuable insights into the genetic architecture of diseases and other phenotypic traits.
This approach has been particularly successful in identifying thousands of variant-trait associations, shedding light on potential genetic factors involved in a wide range of conditions.
However, a significant limitation of GWAS is that most associated variants reside in non-coding regions of the genome, underscoring the critical role of gene regulation in generating diverse phenotypes from identical DNA sequences.

To bridge this gap, various methods have been developed to enhance the interpretation of GWAS findings.
One such approach is MAGMA, which assigns variants to genes based on physical proximity without considering gene expression data.
However, it has been shown that the closest gene to a variant is not necessarily the gene influenced by that variant, revealing the limitations of proximity-based methods.
Transcriptome-Wide Association Studies (TWAS) overcome this limitation by linking genetic variants to gene expression (eQTLs).
TWAS has evolved into a more general framework to test other molecular phenotypes (molQTLs).

Despite their advancements, TWAS and similar approaches typically focus on single variants or individual genes, thereby overlooking the intricate gene--gene interactions that are pivotal according to modern theories such as the omnigenic model of genetic architecture for complex traits.
Capturing these interactions necessitates more sophisticated methodologies.
One effective approach involves leveraging gene modules based on co-expression patterns.
Machine learning methods, such as decomposition techniques and autoencoders, have demonstrated superior performance in identifying these co-expression modules by capturing latent patterns in high-dimensional datasets.
However, the reliance on co-expression alone can lead to false positives, limiting the accuracy of these modules.

To mitigate this, integrating additional data sources is essential.
For instance, PhenoPLIER exemplifies an integrative approach by combining genetic variant-trait associations from GWAS and variant-gene expression-trait associations from TWAS with groups of functionally related genes.
This integration strengthens the reliability of gene modules by cross-validating co-expression patterns with genetic evidence, thus reducing the likelihood of false positives and enhancing the biological relevance of the results.

In this review, we explore how integrating gene modeling machine learning frameworks with genetic and multi-omics data enhances our understanding of the molecular mechanisms underlying complex traits and diseases, effectively approaching genetic studies through the lens of gene networks.

@@

In a simple trait, a single gene can be responsible for the disease (monogenic), for example in the case of Sickle Cell Anemia and Huntington's disease.
However, most traits are not so simplistic and are the result of many mutations across the genome (polygenic).
Even monogenic traits can sometimes be influenced by the polygenic background {[}\protect\hyperlink{ref-VqGYQMXO}{1}{]}.
Genome-wide association studies (GWAS) examine the relationship between specific genetic variants and phenotypes by comparing allele frequencies in individuals of similar ancestry with distinct phenotypic traits (\textbf{Figure \ref{fig:fig1}}).
Both causal and non-causal variants are found to be significant and due to the phenomena of linkage disequilibrium, the results of GWAS are often grouped into risk loci.
These loci represent clusters of variants that demonstrate significant associations, although not every variant may be causal {[}\protect\hyperlink{ref-1873zCaaX}{2}{]}.

Since the first GWAS on age-related macular degeneration (AMD) in 2005, the GWAS catalog rapidly expanded, containing SNP-trait associations across \textgreater5,000 human traits {[}\protect\hyperlink{ref-VRCjjZ9L}{3}{]}.
These studies have successfully identified risk loci for a variety of traits such as type 2 diabetes, auto-immune disease, schizophrenia, major depressive disorder, and many more {[}\protect\hyperlink{ref-17B4AigZT}{4},\protect\hyperlink{ref-yqBcaXHD}{5}{]}.
As GWAS participants increase, more loci are able to be identified.
For example, a study investigating insomnia with a sample size of \textgreater1M identified 202 risk loci {[}\protect\hyperlink{ref-sBTrhvlJ}{6}{]}, compared to earlier studies with a sample size of \textasciitilde110,00, which were only able to identify 3 risk loci {[}\protect\hyperlink{ref-crCAymA1}{7}{]}.
Beyond the identification of risk loci, GWAS have also led to the discovery of novel biological mechanisms in Crohn's disease {[}\protect\hyperlink{ref-s6z5ZuY3}{8},\protect\hyperlink{ref-gxwza7fO}{9}{]}, rheumatoid arthritis {[}\protect\hyperlink{ref-HicyXvA4}{10}{]}, and more.

Although GWAS have identified numerous genetic variants associated with complex traits, translating these findings into biological insights remains challenging {[}\protect\hyperlink{ref-17B4AigZT}{4}{]}.
For example, genes can have epistatic interactions where a secondary loci affects a primary loci.
The peripheral effects of the secondary loci are often too small to be picked up through GWAS alone {[}\protect\hyperlink{ref-IrGVpNji}{11}{]}.
Furthermore, any significant GWAS variants are located in non-coding regions, making it difficult to identify the specific genes that drive trait associations {[}\protect\hyperlink{ref-BqjxxhyF}{12}{]}.
A simple strategy to link GWAS-associated variants with a gene is by physical proximity, which typically selects the closest gene to the most significant SNP at each locus.
Approaches like MAGMA {[}\protect\hyperlink{ref-19XiXgYmd}{13}{]} are based on proximity to compute a gene-trait association.
However, SNPs can have distal effects on the expression of the target gene {[}\protect\hyperlink{ref-9Pr9idng}{14},\protect\hyperlink{ref-OBxYzUMh}{15},\protect\hyperlink{ref-167QL5tMV}{16},\protect\hyperlink{ref-WfdUHVDA}{17}{]}.

Recent approaches combine multi-omics data from various cell types and tissues with GWAS to identify potential mechanisms of SNPs and the associated gene through molecular quantitative trait loci (molQTLs) {[}\protect\hyperlink{ref-QIEm5ZWK}{18}{]} (Figure \ref{fig:fig1}).
For example, an expression quantitative trait loci (eQTL) is a genetic region associated with the expression levels of a nearby or distant gene {[}\protect\hyperlink{ref-1FHKg3JK5}{19}{]}.
eQTL studies estimate that the majority of heritability is explained by the combination of multiple weak \emph{trans}-eQTLs {[}\protect\hyperlink{ref-sGgkLsyy}{20}{]}.

In the next section, we briefly review computational approaches that test different molecular phenotypes via QTLs to link variants with genes.
Both GWAS and QTL studies have been penalized by their large multiple testing burden, causing the need to adopt a high level of significance.
This results in GWAS being underpowered to detect all heritability explained by SNPs {[}\protect\hyperlink{ref-17B4AigZT}{4}{]}.

@@

Transcriptome-wide association studies (TWAS) integrate GWAS with gene expression data (\textbf{Figure \ref{fig:fig1}}) from eQTL analysis, to prioritize genes whose expression across different tissues is influenced by GWAS variants {[}\protect\hyperlink{ref-ReOPt75u}{21},\protect\hyperlink{ref-l6ogswV3}{22}{]}.
By leveraging predicted gene expression levels, TWAS provides a mechanistic link between genetic variants and traits, allowing researchers to move beyond associations with individual SNPs to identify putatively causal genes {[}\protect\hyperlink{ref-l6ogswV3}{22}{]}.
Since TWAS models the genetic regulation of gene expression, the approach allows researchers to impute expression levels in GWAS cohorts where expression data may not be available.
A key advantage of TWAS over GWAS lies in its ability to increase interpretability by providing a gene-trait association: TWAS connects trait-associated SNPs (which are mostly non-coding) to genes, which are biologically functional units.

Several TWAS approaches have been introduced {[}\protect\hyperlink{ref-RmhTH35O}{23}{]}, including PrediXcan {[}\protect\hyperlink{ref-Z8bvDdVq}{24}{]}, FUSION {[}\protect\hyperlink{ref-1D63fEEPb}{25}{]} and TIGAR {[}\protect\hyperlink{ref-19Klmizj8}{26}{]}.
However, all of them implement a similar framework that consists in three steps: 1) model training, 2) gene expression imputation, and 3) gene-trait association.
For example, during 1), PrediXcan, builds one expression prediction model per gene and tissue using penalized linear regression with ElasticNet to model sparse genetic architectures.
These models contain weights for each SNP used as predictor for a gene expression in a tissue.
Given genotype data in a cohort without measured gene expression, during 2), the SNPs weights from the models can be used to impute tissue-specific gene expression for individuals.
During 3), a gene-tissue-trait association is computed by correlating the tissue-specific imputed gene expression with the trait of interest.
Most methods (such as Summary-PrediXcan {[}\protect\hyperlink{ref-vLyTudUB}{27}{]} or S-PrediXcan), however, offer a shortcut by computing a gene-tissue-trait association directly from GWAS summary statistics without the need of individual-level data.
This process, however, requires the user to select a tissue of interest, which might not be straightforward {[}\protect\hyperlink{ref-l6ogswV3}{22}{]}.
To address this limitation, approaches such as MultiXcan {[}\protect\hyperlink{ref-1FFzCXo1s}{28}{]} or UTMOST {[}\protect\hyperlink{ref-93R9hBin}{29}{]} combine information across tissues to generate a gene-trait association.
These multi-tissue approaches are generally more powerful than single-tissue ones, although they do not provide a direction of effect (i.e., whether a higher or lower predicted expression is associated with a higher or lower disease risk).

@@

PWAS complements GWAS and TWAS by aggregating genetic variations in protein-coding regions to assess their combined impact on protein function and phenotype (\textbf{Figure \ref{fig:fig1}}) {[}\protect\hyperlink{ref-1Dsj6bEd3}{30}{]}.
Unlike GWAS, which focuses on individual variants, PWAS evaluates the cumulative effects of multiple coding variants, reducing the scale of multiple testing, detecting complex inheritance patterns, and offering deeper insights into potential disease links.
PWAS employs the FIRM machine learning model to evaluate how variants---such as missense, nonsense, and frameshift mutations---affect protein function by assigning impairment scores {[}\protect\hyperlink{ref-1Dsj6bEd3}{30}{]}.
The aggregate effects of these variants on each gene are then assessed, followed by statistical tests to identify significant associations between gene scores and phenotypes.
For example, PWAS confirmed the association of \emph{MUTYH} with colorectal cancer using UK Biobank data, even when no individual variant reached genome-wide significance in GWAS, highlighting its ability to uncover functional associations missed by other methods.
However, PWAS relies on high-quality proteomic data, may miss non-coding variant associations, and is computationally intensive.

Epigenome-wide association studies (EWAS) encompass methodologies such methylome-wide association studies (MWAS) (\textbf{Figure \ref{fig:fig1}}) and histone acetylome-wide association studies (HAWAS) (\textbf{Figure \ref{fig:fig1}}), based on haQTLs and mQLTs respectability, which focus on specific types of epigenetic modifications to reveal their roles in gene regulation and disease etiology {[}\protect\hyperlink{ref-TPqvAJda}{31}{]}.
MWAS targets DNA methylation patterns, identifying loci where methylation changes are linked to particular traits or diseases.
For instance, a study by Shen et al.~demonstrated a causal relationship between DNA methylation and depression, indicating that epigenetic modifications may mediate genetic risk for psychiatric disorders {[}\protect\hyperlink{ref-1DNxNkLsT}{32}{]}.
HAWAS focuses on histone acetylation modifications, which are crucial for regulating chromatin structure and gene expression.
Del Rosario et al.conducted a HAWAS identifying over 2,000 differentially acetylated loci in immune cells from Mycobacterium tuberculosis-infected individuals, linking these changes to gene expression and potassium channel genes like \emph{KCNJ15} in modulating apoptosis and Mtb clearance.
haQTL analysis further revealed variants associated with immune phenotypes, complementing GWAS findings and enhancing understanding of disease mechanisms {[}\protect\hyperlink{ref-1BYHeZlgB}{33}{]}.
The tissue-specific nature of epigenetic modifications, alongside the capacity to capture cellular plasticity and environmental influences, enhances our insight into the effects of genetic variants across distinct tissues, temporal cellular states, and gene-environment interactions {[}\protect\hyperlink{ref-FeaSnwhX}{34},\protect\hyperlink{ref-KfWCLspO}{35}{]}.

@@

In the early 20th century, a significant debate arose between Mendelian geneticists and biometricians concerning the inheritance of complex traits.
Mendelians emphasized single gene traits with discrete inheritance patterns, while biometricians argued that such models could not account for the continuous phenotypic variation observed in populations.
This conflict highlighted a fundamental question: how can discrete genetic factors produce continuous phenotypic variation? {[}\protect\hyperlink{ref-vpIDZCSa}{36},\protect\hyperlink{ref-15ldVppuv}{37}{]}.
The disagreement was resolved by R.A. Fisher in 1918, who introduced the infinitesimal model, demonstrating that if many genes influence a trait, their combined effects can generate continuous phenotypic distributions {[}\protect\hyperlink{ref-Qxk70FFV}{38}{]}.
Despite the success of Fisher's infinitesimal model, the actual number of genes involved in complex traits and the magnitude of their effects remained uncertain for much of the 20th century.
Translating genotypes into phenotypes proved challenging due to the complexity of the genetic mechanisms driving trait variation.
In this context, GWAS brought a new perspective for identifying specific genetic variants with functional effects on complex traits.
Over a decade of GWAS has led to several unexpected findings:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  The strongest GWAS associations exhibit modest effect sizes on disease risk, collectively accounting for only a minor fraction of the predicted heritability (missing heritability) {[}\protect\hyperlink{ref-RASeYPIy}{39}{]}.
  It is now understood that much of this missing heritability arises from a vast number of common variants with small effects that remain undetected in current sample sizes {[}\protect\hyperlink{ref-1AOy1zxAv}{40}{]}.
\item
  A surprising uniform distribution of these small-effect significant variants exists throughout the genome, including non-coding regions, with a notable concentration in regulatory elements such as enhancers and promoters {[}\protect\hyperlink{ref-mwTa2RUK}{41},\protect\hyperlink{ref-lVJVFaaZ}{42}{]}.
  For instance, it is estimated that 71\%--100\% of 1-MB windows in the genome contribute to heritability for schizophrenia {[}\protect\hyperlink{ref-XvQe1H3A}{43}{]}.
\item
  Trait variation is influenced by the broader genome, not only by genes with direct biological links to specific phenotypes.
  However, the idea that nearly all genomic variants could exert direct additive effects on a given phenotype is biologically unlikely {[}\protect\hyperlink{ref-vpIDZCSa}{36}{]}.
\end{enumerate}

In response to these unexpected findings, Boyle et al.~proposed the omnigenic model {[}\protect\hyperlink{ref-vpIDZCSa}{36}{]}, a conceptual framework in which the genetic architecture of complex traits could be explained by highly interconnected gene regulatory networks.
The omnigenic model reframes the earlier infinitesimal model by distinguishing between ``core'' genes, which have a direct biological role in the disease or trait, and ``peripheral'' genes, which influence core genes and thus indirectly affect the disease or trait through regulatory networks {[}\protect\hyperlink{ref-vpIDZCSa}{36}{]}.
The omnigenic model, unlike the traditional infinitesimal model, offers a mechanistic rationale for complex traits architecture based on molecular and cellular biology, understanding the molecular pathways that connect genetic variation to phenotypic traits (\textbf{Figure \ref{fig:fig2}}).

The key proposals of this model are: 1) nearly all genes expressed in cells relevant to the trait have the potential to impact the regulation of core genes 2) that for typical traits, nearly all heritability is attributable to variation near peripheral genes.
Consequently, while core genes function as the primary drivers of disease, it is the cumulative effects of numerous peripheral gene variants that determine polygenic risk {[}\protect\hyperlink{ref-LXvTZzEA}{44}{]}.
This framework provides a possible explanation for several previously identified problems, including widespread pleiotropy (the ability of a single gene to affect multiple traits), polygenicity (the involvement of many genes in the manifestation of a single trait), the bias effect size where small-effect variants can collectively account for a large portion of heritability, and the uniform distribution of these effects across the genome.

@@

In schizophrenia, the application of the omnigenic model has revealed specific core gene sets associated with the disorder.
Notably, the \emph{TCF4} gene set maintained its significant impact even after excluding SNPs within the \emph{TCF4} gene itself.
This finding suggests that peripheral genes within the set contribute to the development of schizophrenia.
Additionally, the observed systematic floor effect across polygenic scores aligns with the model's prediction that most genes expressed in relevant cells contribute to heritability, thereby highlighting the model's potential in explaining the broad genetic contributions to complex psychiatric conditions {[}\protect\hyperlink{ref-fV6THoBG}{45}{]}.

>>>>

This manuscript
(\href{https://pivlab.github.io/annual_review_of_biomedical_data_science/v/517d43c379923fb79c0ec18f4a2aa10ed2beb727/}{permalink})
was automatically generated
from \href{https://github.com/pivlab/annual_review_of_biomedical_data_science/tree/517d43c379923fb79c0ec18f4a2aa10ed2beb727}{pivlab/annual\_review\_of\_biomedical\_data\_science@517d43c}
on October 10, 2024.

@@

Understanding the genetics of complex traits remains one of the key challenges in modern genomics.
Genome-wide association studies (GWAS) have significantly advanced our knowledge by identifying associations between genetic variants and traits, shedding light on the genetic architecture underlying a variety of diseases and phenotypic conditions.
Since the first GWAS on age-related macular degeneration in 2005, the GWAS catalog has grown to include over 5,000 human traits, encompassing risk loci for conditions such as type 2 diabetes, schizophrenia, and rheumatoid arthritis {[}\protect\hyperlink{ref-HicyXvA4}{3}{]}.
However, most GWAS-identified variants reside in non-coding regions, underscoring the importance of gene regulation in generating phenotypic diversity {[}\protect\hyperlink{ref-BqjxxhyF}{4},\protect\hyperlink{ref-9Pr9idng}{5}{]}.
This gap presents a major challenge in translating GWAS findings into biological insights.

To address this limitation, numerous methods have been developed to link GWAS variants to functional outcomes.
Proximity-based methods, such as MAGMA, have been widely used to link GWAS variants to nearby genes based on physical distance {[}\protect\hyperlink{ref-19XiXgYmd}{6}{]}.
However, these methods may not consider long-range acting SNPs, such as the variants in the \emph{FTO} locus that are associated with obesity, which have been shown to influence the \emph{IRX3} gene through long-range interactions {[}\protect\hyperlink{ref-167QL5tMV}{7}{]}.
While useful, proximity-based approaches may miss causal genes if they are not the closest ones.
Transcriptome-wide association studies (TWAS) have proven effective by linking genetic variants to gene expression, thereby enhancing our ability to identify putatively causal genes for various traits {[}\protect\hyperlink{ref-l6ogswV3}{9}{]}.
TWAS employs expression quantitative trait loci (eQTL) data to prioritize genes that are influenced by GWAS variants, offering a more mechanistic link between genetic variation and phenotypic changes.
Nevertheless, TWAS and similar methods typically focus on individual genes or variants, overlooking the complex interactions among genes that are increasingly recognized as vital to understanding complex traits {[}\protect\hyperlink{ref-vpIDZCSa}{10}{]}.

The omnigenic model, introduced by Boyle et al. {[}\protect\hyperlink{ref-vpIDZCSa}{10}{]}, proposes that the genetic architecture of complex traits involves highly interconnected gene networks.
According to this model, core genes directly contribute to a trait, while peripheral genes modulate these core genes through regulatory networks.
This conceptual shift emphasizes the need for methodologies capable of capturing polygenic and network-based interactions inherent in complex diseases.
Machine learning approaches, particularly those that leverage gene co-expression patterns, are well-suited to this task.
Methods such as non-negative matrix factorization and variational autoencoders (VAEs) have demonstrated the ability to extract biologically meaningful gene modules, which represent groups of genes with coordinated expression under specific conditions {[}\protect\hyperlink{ref-170T6ip47}{12}{]}.

In this review, we explore current approaches based on gene modules to integrate genetic studies with other data types.
We discuss how integrating machine learning-derived gene modules with genetic and multi-omics data enhances our understanding of complex traits and diseases.
We highlight the PhenoPLIER framework {[}\protect\hyperlink{ref-NM3rHx1i}{13}{]}, which integrates gene modules derived from transcriptome data with TWAS and drug-induced transcriptional profiles to uncover disease-relevant molecular mechanisms.
This approach moves beyond single-gene analyses by capturing the broader gene networks that contribute to phenotypic outcomes, offering a more nuanced understanding of the molecular basis of human complex traits and paving the way for more effective, personalized therapeutic strategies.

@@

In a simple trait, a single gene can be responsible for the disease (monogenic), for example in the case of Sickle Cell Anemia and Huntington's disease.
However, most traits are not so simplistic and are the result of many mutations across the genome (polygenic).
Even monogenic traits can sometimes be influenced by the polygenic background {[}\protect\hyperlink{ref-VqGYQMXO}{14}{]}.
Genome-wide association studies (GWAS) examine the relationship between specific genetic variants and phenotypes by comparing allele frequencies in individuals of similar ancestry with distinct phenotypic traits (\textbf{Figure \ref{fig:fig1}}).
Both causal and non-causal variants are found to be significant and due to the phenomena of linkage disequilibrium, the results of GWAS are often grouped into risk loci.
These loci represent clusters of variants that demonstrate significant associations, although not every variant may be causal {[}\protect\hyperlink{ref-1873zCaaX}{15}{]}.

Since the first GWAS on age-related macular degeneration (AMD) in 2005, the GWAS catalog rapidly expanded, containing SNP-trait associations across \textgreater5,000 human traits {[}\protect\hyperlink{ref-VRCjjZ9L}{1}{]}.
These studies have successfully identified risk loci for a variety of traits such as type 2 diabetes, auto-immune disease, schizophrenia, major depressive disorder, and many more {[}\protect\hyperlink{ref-17B4AigZT}{2},\protect\hyperlink{ref-yqBcaXHD}{16}{]}.
As GWAS participants increase, more loci are able to be identified.
For example, a study investigating insomnia with a sample size of \textgreater1M identified 202 risk loci {[}\protect\hyperlink{ref-sBTrhvlJ}{17}{]}, compared to earlier studies with a sample size of \textasciitilde110,00, which were only able to identify 3 risk loci {[}\protect\hyperlink{ref-crCAymA1}{18}{]}.
Beyond the identification of risk loci, GWAS have also led to the discovery of novel biological mechanisms in Crohn's disease {[}\protect\hyperlink{ref-s6z5ZuY3}{19},\protect\hyperlink{ref-gxwza7fO}{20}{]}, rheumatoid arthritis {[}\protect\hyperlink{ref-HicyXvA4}{3}{]}, and more.

Although GWAS have identified numerous genetic variants associated with complex traits, translating these findings into biological insights remains challenging {[}\protect\hyperlink{ref-17B4AigZT}{2}{]}.
For example, genes can have epistatic interactions where a secondary loci affects a primary loci.
The peripheral effects of the secondary loci are often too small to be picked up through GWAS alone {[}\protect\hyperlink{ref-IrGVpNji}{21}{]}.
Furthermore, any significant GWAS variants are located in non-coding regions, making it difficult to identify the specific genes that drive trait associations {[}\protect\hyperlink{ref-BqjxxhyF}{4}{]}.
A simple strategy to link GWAS-associated variants with a gene is by physical proximity, which typically selects the closest gene to the most significant SNP at each locus.
Approaches like MAGMA {[}\protect\hyperlink{ref-19XiXgYmd}{6}{]} are based on proximity to compute a gene-trait association.
However, SNPs can have distal effects on the expression of the target gene {[}\protect\hyperlink{ref-9Pr9idng}{5},\protect\hyperlink{ref-167QL5tMV}{7},\protect\hyperlink{ref-OBxYzUMh}{22},\protect\hyperlink{ref-WfdUHVDA}{23}{]}.

Recent approaches combine multi-omics data from various cell types and tissues with GWAS to identify potential mechanisms of SNPs and the associated gene through molecular quantitative trait loci (molQTLs) {[}\protect\hyperlink{ref-QIEm5ZWK}{24}{]} (Figure \ref{fig:fig1}).
For example, an expression quantitative trait loci (eQTL) is a genetic region associated with the expression levels of a nearby or distant gene {[}\protect\hyperlink{ref-1FHKg3JK5}{25}{]}.
eQTL studies estimate that the majority of heritability is explained by the combination of multiple weak \emph{trans}-eQTLs {[}\protect\hyperlink{ref-sGgkLsyy}{26}{]}.

In the next section, we briefly review computational approaches that test different molecular phenotypes via QTLs to link variants with genes.
Both GWAS and QTL studies have been penalized by their large multiple testing burden, causing the need to adopt a high level of significance.
This results in GWAS being underpowered to detect all heritability explained by SNPs {[}\protect\hyperlink{ref-17B4AigZT}{2}{]}.

@@

Transcriptome-wide association studies (TWAS) integrate GWAS with gene expression data (\textbf{Figure \ref{fig:fig1}}) from eQTL analysis, to prioritize genes whose expression across different tissues is influenced by GWAS variants {[}\protect\hyperlink{ref-ReOPt75u}{8},\protect\hyperlink{ref-l6ogswV3}{9}{]}.
By leveraging predicted gene expression levels, TWAS provides a mechanistic link between genetic variants and traits, allowing researchers to move beyond associations with individual SNPs to identify putatively causal genes {[}\protect\hyperlink{ref-l6ogswV3}{9}{]}.
Since TWAS models the genetic regulation of gene expression, the approach allows researchers to impute expression levels in GWAS cohorts where expression data may not be available.
A key advantage of TWAS over GWAS lies in its ability to increase interpretability by providing a gene-trait association: TWAS connects trait-associated SNPs (which are mostly non-coding) to genes, which are biologically functional units.

Several TWAS approaches have been introduced {[}\protect\hyperlink{ref-RmhTH35O}{27}{]}, including PrediXcan {[}\protect\hyperlink{ref-Z8bvDdVq}{28}{]}, FUSION {[}\protect\hyperlink{ref-1D63fEEPb}{29}{]} and TIGAR {[}\protect\hyperlink{ref-19Klmizj8}{30}{]}.
However, all of them implement a similar framework that consists in three steps: 1) model training, 2) gene expression imputation, and 3) gene-trait association.
For example, during 1), PrediXcan, builds one expression prediction model per gene and tissue using penalized linear regression with ElasticNet to model sparse genetic architectures.
These models contain weights for each SNP used as predictor for a gene expression in a tissue.
Given genotype data in a cohort without measured gene expression, during 2), the SNPs weights from the models can be used to impute tissue-specific gene expression for individuals.
During 3), a gene-tissue-trait association is computed by correlating the tissue-specific imputed gene expression with the trait of interest.
Most methods (such as Summary-PrediXcan {[}\protect\hyperlink{ref-vLyTudUB}{31}{]} or S-PrediXcan), however, offer a shortcut by computing a gene-tissue-trait association directly from GWAS summary statistics without the need of individual-level data.
This process, however, requires the user to select a tissue of interest, which might not be straightforward {[}\protect\hyperlink{ref-l6ogswV3}{9}{]}.
To address this limitation, approaches such as MultiXcan {[}\protect\hyperlink{ref-1FFzCXo1s}{32}{]} or UTMOST {[}\protect\hyperlink{ref-93R9hBin}{33}{]} combine information across tissues to generate a gene-trait association.
These multi-tissue approaches are generally more powerful than single-tissue ones, although they do not provide a direction of effect (i.e., whether a higher or lower predicted expression is associated with a higher or lower disease risk).

@@

PWAS complements GWAS and TWAS by aggregating genetic variations in protein-coding regions to assess their combined impact on protein function and phenotype (\textbf{Figure \ref{fig:fig1}}) {[}\protect\hyperlink{ref-1Dsj6bEd3}{34}{]}.
Unlike GWAS, which focuses on individual variants, PWAS evaluates the cumulative effects of multiple coding variants, reducing the scale of multiple testing, detecting complex inheritance patterns, and offering deeper insights into potential disease links.
PWAS employs the FIRM machine learning model to evaluate how variants---such as missense, nonsense, and frameshift mutations---affect protein function by assigning impairment scores {[}\protect\hyperlink{ref-1Dsj6bEd3}{34}{]}.
The aggregate effects of these variants on each gene are then assessed, followed by statistical tests to identify significant associations between gene scores and phenotypes.
For example, PWAS confirmed the association of \emph{MUTYH} with colorectal cancer using UK Biobank data, even when no individual variant reached genome-wide significance in GWAS, highlighting its ability to uncover functional associations missed by other methods.
However, PWAS relies on high-quality proteomic data, may miss non-coding variant associations, and is computationally intensive.

Epigenome-wide association studies (EWAS) encompass methodologies such methylome-wide association studies (MWAS) (\textbf{Figure \ref{fig:fig1}}) and histone acetylome-wide association studies (HAWAS) (\textbf{Figure \ref{fig:fig1}}), based on haQTLs and mQLTs respectability, which focus on specific types of epigenetic modifications to reveal their roles in gene regulation and disease etiology {[}\protect\hyperlink{ref-TPqvAJda}{35}{]}.
MWAS targets DNA methylation patterns, identifying loci where methylation changes are linked to particular traits or diseases.
For instance, a study by Shen et al.~demonstrated a causal relationship between DNA methylation and depression, indicating that epigenetic modifications may mediate genetic risk for psychiatric disorders {[}\protect\hyperlink{ref-1DNxNkLsT}{36}{]}.
HAWAS focuses on histone acetylation modifications, which are crucial for regulating chromatin structure and gene expression.
Del Rosario et al.conducted a HAWAS identifying over 2,000 differentially acetylated loci in immune cells from Mycobacterium tuberculosis-infected individuals, linking these changes to gene expression and potassium channel genes like \emph{KCNJ15} in modulating apoptosis and Mtb clearance.
haQTL analysis further revealed variants associated with immune phenotypes, complementing GWAS findings and enhancing understanding of disease mechanisms {[}\protect\hyperlink{ref-1BYHeZlgB}{37}{]}.
The tissue-specific nature of epigenetic modifications, alongside the capacity to capture cellular plasticity and environmental influences, enhances our insight into the effects of genetic variants across distinct tissues, temporal cellular states, and gene-environment interactions {[}\protect\hyperlink{ref-FeaSnwhX}{38},\protect\hyperlink{ref-KfWCLspO}{39}{]}.

@@

In the early 20th century, a significant debate arose between Mendelian geneticists and biometricians concerning the inheritance of complex traits.
Mendelians emphasized single gene traits with discrete inheritance patterns, while biometricians argued that such models could not account for the continuous phenotypic variation observed in populations.
This conflict highlighted a fundamental question: how can discrete genetic factors produce continuous phenotypic variation? {[}\protect\hyperlink{ref-vpIDZCSa}{10},\protect\hyperlink{ref-15ldVppuv}{40}{]}.
The disagreement was resolved by R.A. Fisher in 1918, who introduced the infinitesimal model, demonstrating that if many genes influence a trait, their combined effects can generate continuous phenotypic distributions {[}\protect\hyperlink{ref-Qxk70FFV}{41}{]}.
Despite the success of Fisher's infinitesimal model, the actual number of genes involved in complex traits and the magnitude of their effects remained uncertain for much of the 20th century.
Translating genotypes into phenotypes proved challenging due to the complexity of the genetic mechanisms driving trait variation.
In this context, GWAS brought a new perspective for identifying specific genetic variants with functional effects on complex traits.
Over a decade of GWAS has led to several unexpected findings:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  The strongest GWAS associations exhibit modest effect sizes on disease risk, collectively accounting for only a minor fraction of the predicted heritability (missing heritability) {[}\protect\hyperlink{ref-RASeYPIy}{42}{]}.
  It is now understood that much of this missing heritability arises from a vast number of common variants with small effects that remain undetected in current sample sizes {[}\protect\hyperlink{ref-1AOy1zxAv}{43}{]}.
\item
  A surprising uniform distribution of these small-effect significant variants exists throughout the genome, including non-coding regions, with a notable concentration in regulatory elements such as enhancers and promoters {[}\protect\hyperlink{ref-mwTa2RUK}{44},\protect\hyperlink{ref-lVJVFaaZ}{45}{]}.
  For instance, it is estimated that 71\%--100\% of 1-MB windows in the genome contribute to heritability for schizophrenia {[}\protect\hyperlink{ref-XvQe1H3A}{46}{]}.
\item
  Trait variation is influenced by the broader genome, not only by genes with direct biological links to specific phenotypes.
  However, the idea that nearly all genomic variants could exert direct additive effects on a given phenotype is biologically unlikely {[}\protect\hyperlink{ref-vpIDZCSa}{10}{]}.
\end{enumerate}

In response to these unexpected findings, Boyle et al.~proposed the omnigenic model {[}\protect\hyperlink{ref-vpIDZCSa}{10}{]}, a conceptual framework in which the genetic architecture of complex traits could be explained by highly interconnected gene regulatory networks.
The omnigenic model reframes the earlier infinitesimal model by distinguishing between ``core'' genes, which have a direct biological role in the disease or trait, and ``peripheral'' genes, which influence core genes and thus indirectly affect the disease or trait through regulatory networks {[}\protect\hyperlink{ref-vpIDZCSa}{10}{]}.
The omnigenic model, unlike the traditional infinitesimal model, offers a mechanistic rationale for complex traits architecture based on molecular and cellular biology, understanding the molecular pathways that connect genetic variation to phenotypic traits (\textbf{Figure \ref{fig:fig2}}).

The key proposals of this model are: 1) nearly all genes expressed in cells relevant to the trait have the potential to impact the regulation of core genes 2) that for typical traits, nearly all heritability is attributable to variation near peripheral genes.
Consequently, while core genes function as the primary drivers of disease, it is the cumulative effects of numerous peripheral gene variants that determine polygenic risk {[}\protect\hyperlink{ref-LXvTZzEA}{47}{]}.
This framework provides a possible explanation for several previously identified problems, including widespread pleiotropy (the ability of a single gene to affect multiple traits), polygenicity (the involvement of many genes in the manifestation of a single trait), the bias effect size where small-effect variants can collectively account for a large portion of heritability, and the uniform distribution of these effects across the genome.

@@

In schizophrenia, the application of the omnigenic model has revealed specific core gene sets associated with the disorder.
Notably, the \emph{TCF4} gene set maintained its significant impact even after excluding SNPs within the \emph{TCF4} gene itself.
This finding suggests that peripheral genes within the set contribute to the development of schizophrenia.
Additionally, the observed systematic floor effect across polygenic scores aligns with the model's prediction that most genes expressed in relevant cells contribute to heritability, thereby highlighting the model's potential in explaining the broad genetic contributions to complex psychiatric conditions {[}\protect\hyperlink{ref-fV6THoBG}{48}{]}.

Li et al.~(2024) identified a significantly connected subgraph formed by cancer-affected coding genes and ncRNAs by focusing on connectivity as a key topological feature.
This approach highlighted the essential role of ncRNAs in linking fragmented cancer-affected genes, consistent with the omnigenic framework's premise that peripheral genes, including non-coding elements, contribute to complex traits such as cancer.
Importantly, the inclusion of ncRNAs enhanced the identification of cancer-related pathways, indicating that a comprehensive network model encompassing ncRNAs is more effective in characterizing disease relationships than models concentrating solely on coding genes {[}\protect\hyperlink{ref-13VkGrWQR}{49}{]}.
